MSF_Groupwork-NBA_shotlog

### 2016-10-16 MSFGroupwork-NBAHotHand.Rmd
add line for each sections and add analysis explanations

---
title: "MSF_Groupwork-NBA_shotlog" 
author: "Team_9"
output:
  pdf_document: default
  html_document: default
---

@@

@@

##Introduction
###What is your theme or research question?
    a. Hot Hand Theory (Artemis)
    b. Defender Proximity Analysis (Reza)
    c. Shot Clock Pressure  (Alex)
    d. Close Game Effect - Small Margin affecting the scoring rate 
    e. Home Advantage (Alex is working in this point)  
    f. Fatigue Effect (Quarter Heat Rate)  
    g. Shooting Percentage (James)
    h. Fatigue Effect (Number of Rest Day)  
    i. Regression Analysis (Alex)

###Introduction/Summary

The purpose of this report is to analyse a chosen dataset and to produce meaningful results using statistical and mathematical analysis. These analyses have been conducted using RStudio software. This report will focused on analysis of a number of areas, demonstrate how the analysis has been conducted, discusses theory used, demonstrate and evaluate produced results. In addition this report will highlight potential areas of future research.

The chosen dataset for this report is a summary of shots made during NBA season 2015-2016. The following areas of research have been selected:

##Introduction
I   What is your theme or research question?
a. Hot Hand Theory (Artemis)
b. Defender Proximity Analysis (Reza)
c. Shot Clock Pressure  (Alex)
d. Close Game Effect - Small Margin affecting the scoring rate 
e. Home Advantage (Alex is working in this point)  
f. Fatigue Effect (Quarter Heat Rate)  
g. Shooting Percentage (James)
h. Fatigue Effect (Number of Rest Day)  
i. Regression Analysis (Alex)

##Theory
I   What do you expect to find?
a.

##Methods
###Dataset Description:

  Data on shots taken during the NBA 2014-2015 season for regular matches throughout the year. The data comprise of 128,609 rows and 22 columns with 16 MB data size with several value in some column is missing. The column titles are generally self-explanatory. It has the following data:   
    1.  Match ID  
    2.  Match date  
    3.  Which team plays as home and away  
    4.  Match result  
    5.  Final score difference  
    6.  Every shots attempt per game for each player  
    7.  Shot result whether it is success or not  
    8.  Type of shoot attempt (2 points or 3 points)  
    9.  Quarter number data (1 to 4)  
    10. Game clock for each quarter  
    11. Time left for a shot (shot clock)  
    12. Who took the shot  
    13. Shot distance from basket  
    14. Who was the nearest defender  
    15. Distance from the nearest defender

  The dataset is useful for evaluating who the best shooter is, who the best defender is, the hot-hand hypothesis, affect of home playing team, close game effect, shooting accuracy from best players, and defender distance effect on shooting accuracy. Data was taken from NBA's REST API provided by [Kaggle](https://www.kaggle.com/dansbecker/nba-shot-logs) and acquired at 30 September 2016.

@@

    The dataset comprise of comprehensive observational data to start to work with. In general, the dataset only have small percentage of missing value (0.19%) so we do not need to remove or assign many data points to default values. And since the column name is not ambiguous it really helps us to understand dataset from the beginning.

###Dataset Limitations  
  There are several constraints we need to mentioned in order to provide same understandings for this group work analysis as follows:  
    1. The dataset consists only for 1 year NBA season (2014-2015)  
    2. The dataset comprise only with regular match (82 matches per team for both home and away matches)  
    3. The playoff and other matches during the year is not included
    4. The dataset does not include free throw shot and results

@@

### Installing libraries

```{r}
install.packages('reshape')
#install.packages('gridExtra')
```

### Calling libraries
```{r}
library(reshape)
library(ggplot2)
library(gridExtra)
```

###Data clean-up

```{r}
df<- read.csv("shot_logs.csv")  #assigning the dataframe to 'df' variables
attach(df) #telling R that this is the dataframe we'll be working with (eliminating the need to use df$ for variables)

df1<- read.csv("shot_logs1010.csv")

# replacing W/L in column "W" with 1/0 (1 for win, 0 for loss)
df[,"W"] <- as.character(df[,"W"])

for (i in 1:nrow(df)) {
  if (df[i,"W"]=="W") df[i,"W"]<- 1
  if (df[i,"W"]=="L") df[i,"W"]<- 0}

df[,"W"] <- as.numeric(df[,"W"])
```

```{r}
#adding columns for names of 'home' and 'away' teams. In addition, WINNER column tells which team won
df[,"MATCHUP"] <- as.character(df[,"MATCHUP"])
df[,"W"] <- as.character(df[,"W"])

for (i in 1:nrow(df)) {
  if (LOCATION[i] == "H") {df$HOME_TEAM [i] <- unlist(strsplit(df$MATCHUP[i]," "))[5]
  df$AWAY_TEAM [i] <- unlist(strsplit(df$MATCHUP[i]," "))[7]
  if (W[i]=="W") df$WINNER[i] <- "HOME" 
  if (W[i]=="L") df$WINNER[i] <- "AWAY"}
  if (LOCATION[i]=="A") {df$HOME_TEAM [i] <- unlist(strsplit(df$MATCHUP[i]," "))[7] 
  df$AWAY_TEAM [i] <- unlist(strsplit(df$MATCHUP[i]," "))[5]
  if (W[i]=="L") df$WINNER[i] <- "HOME"  
  if (W[i]=="W") df$WINNER[i] <- "AWAY"}}

#adding a column 'DATE'
df1[,"MATCHUP"] <- as.character(df1[,"MATCHUP"])
for (i in 1:nrow(df1)) df1$DATE[i] <- as.character(as.Date(unlist(strsplit(df1$MATCHUP[i]," -"))[1], format= "%B %d, %Y"))

restdays <- cast(df1, DATE + HOME_TEAM + AWAY_TEAM + WINNER ~ FGM, value = 'FGM')

#this code is supposed to produce a list of dates when teams play at HOME in DATES column. for some reason it is not working.

for (i in aa3$TEAM) {
  for (j in 1:nrow(restdays)) {
    aa3$DATES<- as.list(aa3$DATES)
    if (restdays$HOME_TEAM[j]==i) aa3$DATES[aa3$TEAM==i]<- c(aa3$DATES[aa3$TEAM==i], restdays$DATE[j])}}

@@

@@

This area of anlysis focuses on investigation of whether playing at 'home' gives the team advantage comparing to playing 'away'.

Literature review demonstrates that

The NULL Hypothesis is:

```{r}
aa<- cast(df1, GAME_ID1 + HOME_TEAM + AWAY_TEAM + WINNER ~ W)

table(aa$WINNER)
#home wins:
hh<-sum(aa$WINNER=="HOME")
#Away wins:
aw<-sum(aa$WINNER=="AWAY")
#percentage of home team winning: 
round(hh/(hh+aw),2)
round(aw/(hh+aw),2)

@@

```{r}
head(aa)
aa1<- cast(aa, HOME_TEAM  ~ WINNER); names(aa1)[1]<-"TEAM"
aa1$WIN_PCT_HOME<-round(aa1$HOME/(aa1$HOME+aa1$AWAY),2)

aa2<- cast(aa, AWAY_TEAM ~ WINNER); names(aa2)[1]<-"TEAM"
aa2$WIN_PCT_AWAY<-round(aa2$AWAY/(aa2$HOME+aa2$AWAY),2)

aa3<- merge (aa1,aa2, by ="TEAM")

for (i in 1:nrow(aa3)) {if(aa3$WIN_PCT_HOME[i]>aa3$WIN_PCT_AWAY[i]) aa3$HOME_ADV[i]<-"YES" else aa3$HOME_ADV[i]<-"NO"}
head(aa3)

aa3<- aa3[order(aa3$WIN_PCT_HOME,decreasing=TRUE),]
rownames(aa4)<-NULL

aa3$TEAM <- factor(aa3$TEAM, levels = aa3$TEAM[order(-aa3$WIN_PCT_HOME)])

ggplot(aa3,aes(x=TEAM)) + geom_point(aes(y= WIN_PCT_HOME, col= "HOME"))+ geom_point(aes( y= WIN_PCT_AWAY, col= "AWAY"))

@@

#Close Game Effect (Justin)

```{r}
# create new variable to seperate overtime

df[, "overtime"] <- as.numeric()

for (i in 1:nrow(df)) {
  if (df[i, "PERIOD"] >= 5) df[i, "overtime"] <- 1
  if (df[i, "PERIOD"] <= 4) df[i, "overtime"] <- 0
}

# accuracy

@@

for (i in (1:7)){
  df.acc[i, "accuracy"] <- sum(df$PERIOD == i & df$FGM == 1)/sum(df$PERIOD == i)
  df.acc[i, "period"] <- i
  if (df.acc$period[i] <= 4) df.acc[i, "overtime"] <- 0
  if (df.acc$period[i] >= 5) df.acc[i, "overtime"] <- 1
}

t.test(df$FGM[df$PERIOD == 1], df$FGM[df$PERIOD == 2])
t.test(df$FGM[df$PERIOD == 2], df$FGM[df$PERIOD == 3])
t.test(df$FGM[df$PERIOD == 3], df$FGM[df$PERIOD == 4])
t.test(df$FGM[df$PERIOD == 1], df$FGM[df$PERIOD == 3])
t.test(df$FGM[df$PERIOD == 2], df$FGM[df$PERIOD == 4])

# accuracy per player per period

df.period <- cast(df, player_id + player_name + PERIOD ~ FGM)

df.period[, "accuracy"] <- as.numeric()

for (i in 1:nrow(df.period)){
  df.period[i, "accuracy"] <- df.period[i, "1"] / sum(df.period[i, "1"], df.period[i, "0"] ) }

# do analysis for top players per team in regards of accuracy per period

# accuracy for Shot Clock

# dealing with missing values
df[is.na(df)] <- 99

acc <- data.frame(seconds.left = c(0:23))

#accuracy for every second
for (i in (0:23)){
  acc[i, "accuracy"] <- (sum(df$SHOT_CLOCK >= i & df$SHOT_CLOCK < i+1 & df$SHOT_CLOCK != 99 & df$FGM == 1)/           sum(df$SHOT_CLOCK >= i & df$SHOT_CLOCK < i+1 & df$SHOT_CLOCK != 99))}

@@

df.acc.players <- cast(df, player_id + player_name + SHOT_CLOCK ~ FGM)

for (i in (1:nrow(df.acc.players))){
  for (j in (0:23)){
    if (df.acc.players$SHOT_CLOCK[i] >= j & df.acc.players$SHOT_CLOCK[i] < j+1) df.acc.players[i, "seconds.left"] <- j
    if (df.acc.players$SHOT_CLOCK[i] == 99) df.acc.players[i, "seconds.left"] <- 99}
}

df.acc.players[, "acc.per.sec"] <- as.numeric()

for (i in (1:nrow(df.acc.players))){
  df.acc.players[i, "acc.per.sec"] <- sum(df.acc.players$'0'[df.acc.players$seconds.left == df.acc.players$seconds.left[i] & df.acc.players$player_id == df.acc.players$players_id[i]])} ### doesnt work as I would like it

ggplot(df.acc.players, aes(x = seconds.left, y = ))

#Assign accuracy per second to main dataframe
df[, "acc.per.sec"] <- as.numeric()

for (i in 1:nrow(df)){
  for (j in 0:23){
    if (df$SHOT_CLOCK[i] >j & df$SHOT_CLOCK[i] <= j+1 & df$SHOT_CLOCK[i] != 99) df[i, "acc.per.sec"] <- acc[j, "accuracy"]
  }
}

@@

# Shooting Percentage Analysis (James)

```{r}
# Replace "made" and "missed" with 1 and 0 respectively

>>>>

---
title: "MSF_Groupwork-NBA_shotlog"
author: "Team_9"
output:
  html_document: default
  pdf_document: default
---

@@

***

@@

##Introduction

The purpose of this report is to analyse the chosen dataset and to produce meaningful results using statistical and mathematical analysis. These analyses have been conducted using RStudio software. This report will focused on analysis of a number of areas, demonstrate how the analysis has been conducted, discusses theory used, demonstrate and evaluate produced results. In addition, where relevant, this report will highlight potential areas of future research.

The chosen dataset for this report is a summary of shots made during NBA season 2014-2015. The following areas of research have been selected:

**1. Team related**
  a. Home Advantage analysis
  b. Effect of rest on teams' performance

**2. Player related**
  a. Effects on Accuracy 
  b. Shot Clock Pressure
  c. Fatigue Effect (Quarter Accuracy Rate)  
  d. Game result impact by field goal attempts
  e. Defender Proximity Analysis
  f. Shooting Distance vs Shot Accuracy 
  g. Hot Hand Theory

###Dataset Description:

  Data on shots taken during the NBA 2014-2015 season for regular matches throughout the year. The data comprise of 128,609 rows and 22 columns with 16 MB data size with several value in some column missing. The column titles are generally self-explanatory. It has the following data:

Data  contains full set of shot attempts by each team/player during the NBA 2014-2015 season for regular matches throughout the year. The data comprise of 128,609 rows and 22 columns with. The column titles (variables) are as follows:   
    1.  Match ID  
    2.  Match date and names of contestants  
    3.  Location: home and away  
    4.  Match result  
    5.  Final score difference  
    6.  shot number  
    7.  FGM: Shot result: success or miss 
    8.  Type of shot attempt (2 points or 3 points)  
    9.  Quarter number (1 to 4, or higher for overtime)  
    10. Game clock for each quarter when the shot has been made 
    11. Shot clock (time left for a shot)  
    12. Name and ID of a player who took the shot  
    13. Shot distance from basket  
    14. Name and ID of the nearest defender  
    15. Distance from the nearest defender

 The data has been acquired from Kaggle website **(https://www.kaggle.com/dansbecker/nba-shot-logs)** on 30 September 2016.

@@

    The dataset comprises of comprehensive observational data to start to work with. In general, the dataset has only a small percentage of missing value (0.19%) so there was not need to remove or assign many data points to default values. Since the column name is not ambiguous it was really helpful to understand the dataset right from the beginning.

    The dataset comprises of comprehensive observational data which enables efficient analysis without extensive 'clean up' of the data. The dataset only have small percentage of missing value (0.19%), therefore there is no need to remove or assign numerous data points to default values. The fact that initial dataset meant that only a small number of additional columns had to be added (see data clean-up part of the code for details.

###Dataset Limitations  
  There are several limitations of the dataset:  
    1. The dataset consists only 1 year NBA season stats (2014-2015)  
    2. The dataset comprise only the data for regular season matches  
    3. The playoff and other matches during the year are not included
    4. The dataset does not include information on free throw shots
    5. The dataset does not include final scores of the games

***

@@

```{r, eval=FALSE, include=FALSE}
# Installing libraries
install.packages('reshape')
install.packages('gridExtra')

@@

```{r, include=FALSE}
# Calling libraries
library(reshape)
library(ggplot2)
library(gridExtra)
library(plyr)

```{r, include=FALSE}
###Data clean-up

df <- read.csv("shot_logs.csv")  #assigning the dataframe to 'df' variables
 #telling R that this is the dataframe we'll be working with (eliminating the need to use df$ for variables)

df1 <- read.csv("shot_logs1010.csv")

# replacing W/L in column "W" with 1/0 (1 for win, 0 for loss)
df[, "W"] <- as.character(df[, "W"])

for (i in 1:nrow(df)) {
  if (df[i, "W"] == "W") df[i, "W"] <- 1
  if (df[i, "W"] == "L") df[i, "W"] <- 0
}

df[, "W"] <- as.numeric(df[, "W"])
```

```{r, include=FALSE}
#adding columns for names of 'home' and 'away' teams. In addition, WINNER column tells which team won
df[, "MATCHUP"] <- as.character(df[, "MATCHUP"])
df[, "W"] <- as.character(df[, "W"])

for (i in 1:nrow(df)) {
  if (df$LOCATION[i] == "H") {
    df$HOME_TEAM [i] <- unlist(strsplit(df$MATCHUP[i], " "))[5]
    df$AWAY_TEAM [i] <- unlist(strsplit(df$MATCHUP[i], " "))[7]
    if (df$W[i] == "W") df$WINNER[i] <- "HOME" 
    if (df$W[i] == "L") df$WINNER[i] <- "AWAY"
  }
  if (df$LOCATION[i] == "A") {
    df$HOME_TEAM [i] <- unlist(strsplit(df$MATCHUP[i], " "))[7] 
    df$AWAY_TEAM [i] <- unlist(strsplit(df$MATCHUP[i], " "))[5]
    if (df$W[i] == "L") df$WINNER[i] <- "HOME"  
    if (df$W[i] == "W") df$WINNER[i] <- "AWAY"
  }
}

#adding a column 'DATE'
df1[, "MATCHUP"] <- as.character(df1[, "MATCHUP"])
for (i in 1:nrow(df1)) df1$DATE[i] <- as.character(as.Date(unlist(strsplit(df1$MATCHUP[i], " -"))[1], format = "%B %d, %Y"))

#creating a column for 'TEAM'. This column relates to the team whose player made a shot described by the dataframe
for (i in 1:nrow(df1)) df1$TEAM[i] <-unlist(strsplit(df1$MATCHUP[i], " "))[5]

#preparing dataframe for rest days calculation
df1$BLANK2<-""
restdays2 <- cast(df1, TEAM + DATE + LOCATION + W + FINAL_MARGIN ~ FGM)
restdays2$DATE<- as.Date(restdays2$DATE)

***

@@

This area of analysis focuses on investigation of whether playing at 'home' gives the team advantage comparing to when playing 'away'. Esssentially, this analysis focuses on effect of the game lovation (home/away) on winning ratio.

The NULL Hypothesis: Teams playing at home is more likely to win then when playing away.

###Methods
Steps to test the hypothesis are as follows:  
    1. Calculate a total number of games, which has been won by a home or away team
    2. Compare the total win numbers to identify whether home advantage is evident using means of the variables
    3. Extend the analysis to review of the home/away wins on a team level 
    4. Plot the team level data 
    5. Discuss the findings of the analysis

```{r, include=FALSE}
#preparing a dataframe for home advantage analysis
aa <- cast(df1, GAME_ID1 + HOME_TEAM + AWAY_TEAM + WINNER ~ W)

#table showing a number of games won by home and away teams in the whole season
table(aa$WINNER)

#home wins:
hh <- sum(aa$WINNER == "HOME")
#Away wins:
aw <- sum(aa$WINNER == "AWAY")

#percentage of home team winning: 
round(hh / (hh + aw), 4)
round(aw / (hh + aw), 4)

@@

```{r, include=FALSE}
#preparing a dataframe for analysis of games when a team played at home 
aa1 <- cast(aa, HOME_TEAM ~ WINNER); names(aa1)[1] <- "TEAM"
aa1$WIN_PCT_HOME <- round(aa1$HOME / (aa1$HOME + aa1$AWAY), 2)

#preparing a dataframe for analysis of games when a team played at away 
aa2 <- cast(aa, AWAY_TEAM ~ WINNER); names(aa2)[1] <- "TEAM"
aa2$WIN_PCT_AWAY <- round(aa2$AWAY / (aa2$HOME + aa2$AWAY), 2)

#merging two dataframes together to enable the effective analysis of win ratios for teams playing at home and away.
aa3 <- merge (aa1, aa2, by ="TEAM")

#ordering the dataframe by win percentage at home.
aa3 <- aa3[order(aa3$WIN_PCT_HOME, decreasing = TRUE),]

#ordering factors of the dataframe by win percentage at home.
aa3$TEAM <- factor(aa3$TEAM, levels = aa3$TEAM[order(-aa3$WIN_PCT_HOME)])

#creating a column which shows a difference between home and away winning percentage rate
aa3$WIN_PCT_DIFF<- aa3$WIN_PCT_HOME - aa3$WIN_PCT_AWAY

@@

###Findings
Table 1: Proportion of total games split by Location win factor

X     | HOME WIN | AWAY WIN
----- | -------- | ---
Game  | 506      | 398
Perc  | 55.97%   | 44.03%

Tables 2 & 3: Winning ratio of teams playing at home vs games playing away

TEAM        | GSW | ATL | POR | MEM | SAS | CLE | HOU | LAC | OKC | DAL | NOP | WAS | TOR | CHI | MIL  
------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----
Win % HOME  |   93| 88  | 81  | 77  | 75  | 72  | 72  | 72  | 71  | 68  | 68  | 68  | 66  | 59  | 59 
Win % AWAY  |   69| 71  | 54  | 67  | 50  | 53  | 61  | 57  | 42  | 61  | 40  | 43  | 59  | 66  | 45 
Win % Diff  |   24| 17  | 27  | 10  | 25  | 19  | 11  | 15  | 29  | 7   | 28  | 25  | 7   | -7  | 14

TEAM        | PHX | IND | BOS | CHA | UTA | DEN | BKN | SAC | MIA | DET | LAL | ORL | PHI | NYK |MI 
------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----
Win % HOME  | 57  | 52  | 48  | 47  | 45  | 42  | 41  | 41  | 39  | 38  | 31  | 30  | 28  | 27  | 26 
Win % AWAY  | 47  | 35  | 32  | 41  | 35  | 30  | 44  | 30  | 48  | 39  | 23  | 31  | 16  | 14  | 18 
Win % Diff  | 10  | 17  | 16  | 6   | 10  | 12  |-3   | 11  |-9   |-1   |    8|   -1|   12|   13| 8

```{r}
#plot showing home and away win ratio per team
ggplot(aa3, aes(x = TEAM)) + geom_point(aes(y = WIN_PCT_HOME, col = "HOME")) + geom_point(aes(y = WIN_PCT_AWAY, col = "AWAY")) 
```

The plot shows that out of 30 teams:
  25 teams have won more games at home than away
  3 teams have won more games away than at home
  2 teams home and away win ratios are almost equal (difference of less than or equal to 1%)

###Variance test Analysis
```{r}

```
**JUSTING TO SEE IF THERE IS ANYTHING WE CAN INCLUDE HERE

###regression analysis

```{r, include=FALSE}
margreg<- lm(FINAL_MARGIN ~ LOCATION, restdays2)
summary(margreg)
```

Table 4: Regression Analysis

  X         |Estimate |Std. Error| t value| Pr(>t)    
------------|---------|--|--|------------------------
(Intercept) |-2.12919 |   0.51589 | -4.127 |3.84e-05 ***
LOCATIONH   |4.28556  |  0.63881  | 6.709 |2.63e-11 ***
rest        |-0.01129 |   0.25992 | -0.043 |   0.965

Adjusted R-squared = 0.02392

###Discussion

The analysis shown that there is a clear evidence that a team playing at home is more likely to win with an average win probability of 55.97%. The same conclusion is evident from the analysis conducted on a team level where 25 teams are winning more games at home than away. The result shown that the NULL hypothesis should be rejected as a team playing at home is more likely to win then an away team. Regression analysis shown that playing at home gives teams on average additional 4.29 margin points.

The result of this research also confirms recent trend of decline in home advantage over the years: probability of a home team winning has reduced from around 65% in 1975-1992 to average of 60.3% in 1993-2011 and to around 58.5% for 2011-15. Home win ratio was 57.4% win in 2014-2015 season. The analysis demonstrated that the home win ratio went down even further to 56.0% in 2015-2016 season. Based on the literature review this trend could be potentially explained by the following factors (Economist, 2015):

  1. Improvement of travel conditions. This means that when playing away, teams are now more likely to stay at better hotels and fly charterd planes more often than in the past. These changes result in teams being less tired as a consequence of travel and being better physically prepared for the games.

  2. Change in style of play in NBA. The game now is more open with less physical contact, which reduces number of fouls and potential impact of 'notorious home-team bias' by referees (Economist, 2015).

##Effect of rest on teams' performance

###Introduction

This area of analysis focuses on investigation of whether a number of rest days between games affect performance and more importantly winning ratio of a team.

The Null Hypothesis is: By having longer rest a team is likely to have a higher winning ratio.

###Methods
The steps to calculate this measurement are as follows:  
    1. Calculating number of rest days between games for all teams
    2. Calculating ratios for effects of rest days on win/loss
    3. Calculating ratios for effects of rest days on win/loss form home/awayCalculating ratios for effects of rest days on win/loss
    4. Discuss the findings of the data

```{r, include=FALSE}

#calulating number of rest days between teams
for (i in 2: nrow (restdays2)) {
  if (restdays2$TEAM[i]==restdays2$TEAM[i-1])
    restdays2$rest[i]<- as.numeric(restdays2$DATE[i] - restdays2$DATE[i-1]-1)
  else 
    (restdays2$rest[i]<- "NA")  
}
restdays2$BLANK<-""

restdays2$rest <- as.numeric (restdays2$rest)

#regression analysis
margreg<- lm(FINAL_MARGIN ~ LOCATION + rest, restdays2)
summary(margreg)

#creating dataframe for total rest days
restdays4<- cast (restdays2, rest ~  W)
colnames(restdays4)<- c( "rest", "L","W")

#creating a column for total restdays for a team when lost and won to enable calculation of ratios
restdays4$total<-restdays4$L + restdays4$W
restdays4$dist <-   round (restdays4$total * 100 /sum (restdays4$total) ,2)

#plotting distribution of rest days to establish which of the rows should be analysed
restdays4[,c(1,4,5)]
#The distribution of rest days show that 92% of games are covered by 0,1 or 2 days of rest and therefore further analysis will only focus on these 3 sections

restdays4$ L.pct <-   round(restdays4$L*100/(restdays4$L+restdays4$W),2) 
restdays4$ W.pct <-   round(restdays4$W*100/(restdays4$L+restdays4$W),2)

print (restdays4[1:3,c(1,3,7)])

#creating a new dataframe for summary of home/away win losses based on number of rest days.
restdays3<- cast (restdays2, rest ~ LOCATION + W) ;

#renaming the columns
colnames(restdays3)<- c( "rest", "AL","AW","HL", "HW")

#caluclating ratios
restdays3$ AL.pct <-   round(restdays3$AL*100/(restdays3$AL+restdays3$AW),2) 
restdays3$ AW.pct <-   round(restdays3$AW*100/(restdays3$AL+restdays3$AW),2)
restdays3$ HL.pct <-   round(restdays3$HL*100/(restdays3$HL+restdays3$HW),2)
restdays3$ HW.pct <-   round(restdays3$HW*100/(restdays3$HL+restdays3$HW),2)

print (restdays3[1:3,c(1,7,9)])

#Testing of rest data

#t.test(x=restdays2$W[restdays2$rest==0], y= restdays2$W[restdays2$rest==1])
#chisq.test(x=restdays2$W[restdays2$rest==0], y= restdays2$W[restdays2$rest==1])

#restdays2$W[restdays2$rest==0,]

###Findings

Table 5: Distribution of number of games per rest days

 Rest Days| Games | Distribution
----------|-------|---
    0  | 427 | 23.62
    1  | 990 | 54.76
    2  | 254 | 14.05
    3  |  59 |  3.26
    4  |  13 |  0.72
    5  |   3 | 0.17
    7  |   6 | 0.33
    8  |  20 | 1.11
    9  |   3 | 0.17
    10 |   3 | 0.17
   1st game of season  |  30|  1.66

The distribution of rest days shows that 92% of games are covered by 0,1 or 2 days of rest and therefore further analysis will only focus on these 3 sections.

Table 6: Win Rate split by number of rest days between games

  rest days| number of wins| win percentage
-|-|-
    0| 201| 47.07
    1| 496| 50.10
    2| 137| 53.94

Table 7: Win Rate split by number of rest days between games and location

  rest days| home wins percentage | away wins percentage|
-|-|-
    0|   53.54 |44.33 
    1 | 54.68 |  44.24
    2  | 59.57 | 46.90

###Discussion

The analysis demonstrated that overall winning ratio improves with increase in  rest days: back-to-back games had 47.07% winning ratio, 1 day rest increased winning ratio to 50.10% and with 2 days rest the winning ration has increased even further to 53.94%.

By extending the analysis to the team level, the results have demonstrated similar tendency for home  games where winning ratio has increased from 53.54% to 54.68% and 59.57% respectively with 0, 1 and 2 days of rest. Somewhat interestingly the analysis of effect of rest days on winning ratio of away games demonstrated that there is hardly any difference between 0 and 1 days of rest (44.33% and 44.24%). However with 2 days of rest teams' average winning ratio have increased to 46.90% (2.66% increase). This perhaps could be explained by the fact that teams were able to minimise effect of travelling when having a bigger window between games.

Based on the above, the analysis shown that the null hypothesis should be rejected as there is a clear evidence that the higher rest days correspond to higher winning ratio

###Further research

To better understand effects of rest and home/away advantage the future research should focus on distances the teams have to travel between the games. For example if a team is travelling from one coast to another- it is likely to affect them more comparing to a team from New York 'travelling' to play an away gamein New Jersey.

***

##Effects on accuracy

The potential effects of fatigue and stress on the accuracy of shots are investigated in this part of the analysis.

Fatigue in basketball is a broadly analysed topic in physiological research. Latest literature suggests that elite basketball players cope well with the effects of fatigue so it does not affect their shooting kinematics. This results in no difference for their shot accuracy even if the players are fatigued (Erculj and Supej, 2009). This research paper analyses whether these previous findings can be confirmed for the NBA season 2014-2015.

Regarding the effect of psychological stress on shot accuracy, previous research is not as decisive as for the effects of fatigue. While Ahart(1973, in P.C. Kendall & S.D Hollon (1979) *Cognitive-Behavioral Interventions: Theory, Research, and Procedures*) concluded that stressful events like small or big margins for the game score have an effect on players free-throw accuracy, latest research introducing more controlled conditions states that there is no effect of stress on players shot accuracy (Mascret et al., 2016). The indecisiveness of previous literature led to a deeper analysis in this research paper.

Hypotheses for this area of analysis are:

Fatigue: The Null hypothesis states that there is no significant difference (p < .05) for the shot accuracy in later periods of the game.

Stress: The Null hypothesis states that there is no significant difference (p < .05) for the shot accuracy in more stressful events.

###Methods

In this analysis fatigue was measured as the period of the game, as it can be assumed that the later in the game the higher the fatigue of the players.

The steps to test the first hypothesis were as follows:
  1. The overall accuracy for each period was calculated, based on the number of shots made (FGM = 1) and the number of shots taken for the specific period (FGM = 1 & FGM = 0)
  2. This results in a percentage value of shots made for each of the periods
  3. A variable called "Overtime" was introduced to distinguish between periods in the regular game time and overtime
  4. The average accuracy for regular time and overtime was calculated
  5. A T-test was used to examine whether or not these values were significant different at a level of significance of 5%.

Psychological stress was measured through the remaining time on the shot clock for each shot taken. It was assumed that a player would experience more stress the lower the shot clock goes.

The steps to test the second hypothesis were as follows:
  1. As the shot clock variable contained 5567 missing values, those were excluded from the analysis
  2. To simplify the shot clock variable, decimals were removed from the data so every shot taken was assigned to 0 to 24 seconds left on the shot clock
  3. The overall accuracy for each second was calculated using the FGM variable

```{r, include=FALSE}

### FATIGUe

# accuracy per period

@@

for (i in (1:7)){
  df.acc[i, "accuracy"] <- sum(df$PERIOD == i & df$FGM == 1) / sum(df$PERIOD == i)
  df.acc[i, "period"] <- i
  if (df.acc$period[i] <= 4) df.acc[i, "overtime"] <- 0
  if (df.acc$period[i] >= 5) df.acc[i, "overtime"] <- 1
}

regular <- sum(df.acc$accuracy[df.acc$period <= 4])/4
overtime <- sum(df.acc$accuracy[df.acc$period >=5])/3

t.test(df$FGM[df$PERIOD == 1], df$FGM[df$PERIOD == 2])
t.test(df$FGM[df$PERIOD == 2], df$FGM[df$PERIOD == 3])
t.test(df$FGM[df$PERIOD == 3], df$FGM[df$PERIOD == 4])
t.test(df$FGM[df$PERIOD == 1], df$FGM[df$PERIOD == 3])
t.test(df$FGM[df$PERIOD == 2], df$FGM[df$PERIOD == 4])
t.test(df$FGM[df$PERIOD == c(1:4)], df$FGM[df$PERIOD == c(5:7)])

### STRESS
# accuracy for Shot Clock

# dealing with missing values
df1[is.na(df1)] <- 99

acc <- data.frame(seconds.left = c(0:24))

#accuracy for every second
for (i in (0:24)){
  acc[i, "accuracy"] <- (sum(df1$SHOT_CLOCK >= i & df1$SHOT_CLOCK < i + 1 & df1$SHOT_CLOCK != 99 & df1$FGM == 1) / sum(df1$SHOT_CLOCK >= i & df1$SHOT_CLOCK < i + 1 & df1$SHOT_CLOCK != 99))
}

@@

df.acc.players <- cast(df1, player_id + player_name + SHOT_CLOCK ~ FGM)

for (i in (1:nrow(df.acc.players))){
  for (j in (0:24)){
    if (df.acc.players$SHOT_CLOCK[i] >= j & df.acc.players$SHOT_CLOCK[i] < j + 1) df.acc.players[i, "seconds.left"] <- j
    if (df.acc.players$SHOT_CLOCK[i] == 99) df.acc.players[i, "seconds.left"] <- 99
  }
}

df.acc.secondsplayers<- df.acc.players

df.acc.secondsplayers$SHOT_CLOCK <- NULL

colnames(df.acc.secondsplayers)<- c( "player_id", "player_name","X0", "X1","seconds.left")

df.acc.secondsplayers2 <- ddply(df.acc.secondsplayers, c("player_id", "player_name","seconds.left"), summarise,
X0= sum(X0),
X1= sum(X1))

df.acc.secondsplayers2$acc <-  df.acc.secondsplayers2$X1/(df.acc.secondsplayers2$X0+df.acc.secondsplayers2$X1)

df.acc.secondsplayers2[df.acc.secondsplayers2 == 99] <- NA

df.acc.secondsplayers3 <- data.frame(row.names = c(0:24))
df.acc.secondsplayers3$seconds.left <- c(0:24)

for (i in df.acc.secondsplayers3$seconds.left){
  df.acc.secondsplayers3$acc.top[df.acc.secondsplayers3$seconds.left == i] <- sum(df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "james harden"][!is.na(df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "james harden"])], df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "lebron james"][!is.na(df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "lebron james"])], df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "russell westbrook"][!is.na(df.acc.secondsplayers2$acc[df.acc.secondsplayers2$seconds.left == i & df.acc.secondsplayers2$player_name == "russell westbrook"])])/3}

ggplot(NULL) + geom_smooth(data = df.acc.secondsplayers3, aes(x = seconds.left, y = acc.top, color = "Top Players")) + geom_smooth(data = acc, aes(x = seconds.left, y = accuracy, color = "Average"))

t.test(acc$accuracy, df.acc.secondsplayers3$acc.top)

@@

###Results

The accuracies for each period are detailed in the table below.

Period  | Accuracy
------- | ---------
1       | 46%
2       | 45%
3       | 46%
4       | 44%
5       | 39%
6       | 43%
7       | 37%
regular | 45%
overtime | 40%

The T-tests for period 1 and period 2 was significant, t(65301) = 2.42, p = .015, as well as the T-test for period 3 and 4, t(60757) = 4.24, p < .01. Further the T-test for period 2 and 3 was not significant, t(63843) = -1.53, p = .126, as well as the T-test for period 1 and 3, t(65989) = 0.87, p = .38. 
The T-test for regular time and overtime was highly significant, t(383) = 2.77, p < .01.

The changes in the overall accuracy per second are shown in the graph below.

```{r}
graph.acc <- ggplot(acc, aes(x = seconds.left, y = accuracy)) + geom_point() + geom_smooth()

graph.acc + labs(title = "Overall Accuracy per second left on shot clock") + xlab("Seconds left on shot clock") + ylab("Overall Accuracy")
```

### Discussion
